---
title: "New phase I methods in escalation"
subtitle: "Introducing new methods in `escalation` for TPI, mTPI and Neuenschwander et al.'s design for phase I trials, plus MCMC CRM methods."
summary: "Introducing new methods in `escalation` for TPI, mTPI and Neuenschwander et al.'s design for phase I trials, plus MCMC-based CRM methods."
authors: 
  - admin
categories:
- Code
tags:
  - escalation
  - R
  - DoseFinding
projects: []
date: '2020-10-20'
lastmod: '2020-10-20T17:07:58Z'
draft: false
featured: no
image:
  caption: 'Photo by [Samuel Isaac on Unsplash](https://unsplash.com/photos/83AdeG3iPMk)'
  focal_point: ''
  preview_only: no
bibliography: library.bib
---

## Introduction

This week I updated the [`escalation`](https://CRAN.R-project.org/package=escalation) [@escalation] package on CRAN with new methods to implement several phase I and phase I/II dose-finding designs.
This post describes the addition of the phase I methods TPI, mTPI, and the logistic model method of @Neuenschwander2008, and also the addition of MCMC-based CRM methods.

## TPI
The toxicity probability interval (TPI) design was introduced by @Ji2007.
It is one of a series of dose-finding trial designs that works by partitioning the probability of toxicity into a set of intervals.
These designs make dose-selection decisions that are determined by the interval in which the probability of toxicity for the current dose is believed to reside.

Core to this design is a beta-binomial Bayesian conjugate model.
For hyperparameters $\alpha$ and $\beta$, let the probability of toxicity at dose $i$ be $p_i$, with prior distribution

$$p_i \sim Beta(\alpha, \beta).$$

If $n_i$ patients have been treated at dose $i$, yielding $x_i$ toxicity events, the posterior distribution is 

$$ p_i | data \sim Beta(\alpha + x_{i}, \beta + n_{i} - x_{i}).$$

Using this distribution, let the standard deviation of $p_i$ be denoted by $\sigma_i$.
The design seeks a dose with probability of toxicity close to some pre-specified target level, $p_T$.
The entire range of possible values for $p_i$ can be broken up into the following intervals: 

* The underdosing interval (UI), defined as $(0, p_{T} - K_{2} \sigma_{i})$;
* The equivalence interval (EI), defined as $(p_{T} - K_{2} \sigma_{i}, p_{T} + K_{1} \sigma_{i})$;
* The overdosing interval (OI), defined as $(p_{T} + K_{1} \sigma_{i}, 1)$;

for pre-specified model constants, $K_1, K_2$.
These intervals are  mutally-exclusive and mutually-exhaustive, meaning that every possible probability belongs to precisely one of them.
In other words, these intervals form a partition of the probability space, $(0, 1)$.

Using the posterior distribution, we can calculate the three probabilities

$$p_{UI} = Pr(p_i \in \text{UI}), \enspace p_{EI} = Pr(p_i \in \text{EI}), \enspace p_{OI} = Pr(p_i \in \text{OI}).$$

By definition, $p_{UI} + p_{EI} + p_{OI} = 1$.
The logical action in the dose-finding trial depends on which of these three probabilities is the greatest. 
If $p_{UI} > p_{EI}, p_{OI}$, then the current dose is likely an underdose, so our desire should be to escalate dose to $i+1$.
In contrast, if $p_{OI} > p_{UI}, p_{EI}$, then the current dose is likely an overdose and we will want to de-escalate dose to $i-1$ for the next patient.
If $p_{EI} > p_{UI}, p_{OI}$, then the current dose is deemed sufficiently close to $p_T$ and we will want to stay at dose-level $i$.

Further to these rules regarding dose-selection, the following rule is used to avoid recommending dangerous doses.
A dose is deemed inadmissible for being excessively toxic if

$$ Pr(p_{i} > p_{T} | data) > \xi,$$

for a certainty threshold, $\xi$.
If a dose is excluded by this rule, it should not be recommended by the model.
Irrespective the probabilities $p_{UI}, p_{EI}, p_{OI}$, the design will recommend to stay at dose $i$ rather than escalate to a dose previously identified as being inadmissible.
Furthermore, the design will advocate stopping if the lowest dose is inferred to be inadmissible.

In their paper, the authors demonstrate acceptable operating performance using $\alpha = \beta = 0.005$, $K_{1} = 1$, $K_{2} = 1.5$ and $\xi = 0.95$.
See @Ji2007 for full details.


### Implementation in `escalation`

Let us specify the model that the authors investigate in their publication:

```{r, message=FALSE}
library(escalation)

model <- get_tpi(num_doses = 5, target = 0.3, alpha = 0.005, beta = 0.005, 
                 k1 = 1, k2 = 1.5, exclusion_certainty = 0.95)
```

As with all models in `escalation`, we fit the model to some outcomes using code like:

```{r}
fit <- model %>% fit('1NNT') 
```

and learn the dose recommended for the next patient(s):

```{r}
fit %>% recommended_dose()
```

In their Table 1, @Ji2007 recommend some model choices based on notional cohort outcomes.
We can reproduce their findings using, for example:

```{r}
model %>% 
  get_dose_paths(cohort_sizes = 6, next_dose = 2) %>% 
  graph_paths()
```

I.e. if there are six patients at the current dose, this particular parameterisation will only advocate escalation if exactly zero toxicities are seen.
If one or two toxicities are seen, the design advocates staying at dose 2, otherwise deescalation is recommended.

## mTPI
The modified toxicity probability interval (mTPI) design was introduced by @Ji2010.
As the name suggests, it is a modification of the earlier TPI design.
mTPI is very similar to TPI, again using a method that partitions the probability of toxicity into a set of intervals.

This design uses a beta-binomial Bayesian conjugate model.
For hyperparameters $\alpha$ and $\beta$, let the probability of toxicity at dose $i$ be $p_i$, with prior distribution

$$p_i \sim Beta(\alpha, \beta).$$

If $n_i$ patients have been treated at dose $i$, yielding $x_i$ toxicity events, the posterior distribution is 

$$ p_i | data \sim Beta(\alpha + x_{i}, \beta + n_{i} - x_{i}).$$

The design seeks a dose with probability of toxicity close to some pre-specified target level, $p_T$.
The entire range of possible values for $p_i$ can be broken up into the following intervals: 

* The underdosing interval (UI), defined as $(0, p_{T} - \epsilon_{1})$;
* The equivalence interval (EI), defined as $(p_{T} - \epsilon_{1}, p_{T} + \epsilon_{2})$;
* The overdosing interval (OI), defined as $(p_{T} + \epsilon_{2}, 1)$;

for pre-specified model constants, $\epsilon_{1}, \epsilon_{2}$.
These intervals are  mutally-exclusive and mutually-exhaustive, meaning that every possible probability belongs to precisely one of them.
In other words, these intervals form a partition of the probability space, $(0, 1)$.

For a continuous random variable $X$ with cumulative probability mass function $F(x)$ (i.e. $Pr(X < x) = F(x)$), the authors define the unit probability mass (UPM) for an interval $(a, b)$ to be $(F(b) - F(a)) / (b - a)$.
That is, the UPM is the probability mass in an interval divided by the width of the interval, and can be interpreted as the average probability density of the interval.

Then, using the posterior distribution identified above, we calculate the three UPMs

$$UPM_{UI} = Pr(p_i \in \text{UI}) / (p_{T} - \epsilon_{1}),$$
$$UPM_{EI} = Pr(p_i \in \text{EI}) / (\epsilon_{1} + \epsilon_{2}),$$

and 

$$UPM_{OI} = Pr(p_i \in \text{OI}) / (1 - p_{T} + \epsilon_{2}).$$

The logical action in the dose-finding trial depends on which of these three quantities is the greatest. 
If $UPM_{UI} > UPM_{EI}, UPM_{OI}$, then the current dose is likely an underdose, so our desire should be to escalate dose to $i+1$.
In contrast, if $UPM_{OI} > UPM_{UI}, UPM_{EI}$, then the current dose is likely an overdose and we will want to de-escalate dose to $i-1$ for the next patient.
If $UPM_{EI} > UPM_{UI}, UPM_{OI}$, then the current dose is deemed sufficiently close to $p_T$ and we will want to stay at dose-level $i$.

Further to these rules regarding dose-selection, the following rule is used to avoid recommending dangerous doses.
A dose is deemed inadmissible for being excessively toxic if

$$ Pr(p_{i} > p_{T} | data) > \xi,$$

for a certainty threshold, $\xi$.
If a dose is excluded by this rule, it should not be recommended by the model.
Irrespective the values of $UPM_{UI}, UPM_{EI}$ and $UPM_{OI}$, the design will recommend to stay at dose $i$ rather than escalate to a dose previously identified as being inadmissible.
Furthermore, the design will advocate stopping if the lowest dose is inferred to be inadmissible.

In their paper, the authors demonstrate acceptable operating performance using $\alpha = \beta = 1$, $\epsilon_{1} = 0.05$, $\epsilon_{2} = 0.05$ and $\xi = 0.95$.
See @Ji2010 for full details.


### Implementation in `escalation`

Again, let us investigate one of the parameterisations that the authors investigate in their publication:

```{r, message=FALSE}
library(escalation)

model <- get_mtpi(num_doses = 5, target = 0.3, alpha = 1, beta = 1, 
                  epsilon1 = 0.05, epsilon2 = 0.05, exclusion_certainty = 0.95)
```

As before, we fit the model and learn the dose recommended for the next patient(s) using:

```{r}
fit <- model %>% fit('1NNT') 
fit %>% recommended_dose()
```

In their Figure 2, the authors again enumerate model advice over a range of cohort outputs.
Just as we did with TPI, enumerating the dose recommendations after a cohort of 6 has been treated at dose 2:

```{r}
model %>% 
  get_dose_paths(cohort_sizes = 6, next_dose = 2) %>% 
  graph_paths()
```

We see that after `2NNNNNT`, the mTPI advocates escalation whereas TPI advocated sticking at dose 2.
The advice from the two models matches in all other outcomes.

## Neuenschwander _et al._
@Neuenschwander2008 (NBG) introduced a derivative of the CRM for dose-escalation clinical trials using the model:

$$ \text{logit} p_i = \alpha + \exp{(\beta)} \log{(x_i / d^*)}, $$

where $p_i$ is the probability of toxicity at the $i$th dose, $x_i$, and $d^*$ is a reference dose.
Here $\alpha$ and $\beta$ are model parameters on which the authors place a bivariate normal prior. 
This model is very similar to the two-parameter logistic CRM.
However, a notable difference is that the dose, $x_i$, enters the model as a covariate.
This dispenses with the toxicity skeleton that is used in the CRM.


### Implementation in `escalation`

The heavy lifting required to fit the model is performed by `trialr` and `rstan`.
`escalation` merely composes the model fit in such a way that it can be used with the other dose-selection modules.

For illustration, let us reproduce the analysis in @Neuenschwander2008 that the authors used to demonstrate the flexibility of a two-parameter approach.
In a trial of 15 doses, the investigators saw outcomes:

```{r}
library(escalation)

dose <- c(1, 2.5, 5, 10, 15, 20, 25, 30, 40, 50, 75, 100, 150, 200, 250)
outcomes <- '1NNN 2NNNN 3NNNN 4NNNN 7TT'
```

Creating a dose-escalation model with NBG's parameters:
```{r}
model <- get_trialr_nbg(real_doses = dose, d_star = 250, target = 0.3,
                        alpha_mean = 2.15, alpha_sd = 0.84,
                        beta_mean = 0.52, beta_sd = 0.8,
                        seed = 2020)
```

and fitting the model to the observed outcomes:
```{r}
fit <- model %>% fit(outcomes)
fit
```

we see that dose 7 is selected for the next cohort using the metric of selecting the dose with posterior expected probability of toxicity closest to the target.
In the above output, `mean_prob_tox` broadly matches the values plotted in the lower right panel of Figure 1 in @Neuenschwander2008.

### Further work
There are a few minor shortcomings of the NBG implementation in `escalation` & `trialr`.
Firstly, NBG propose a bivariate normal prior distribution on $\alpha$ and $\beta$.
However, the implementation in `trialr` currently uses independent normal priors.
Hopefully, this will be addressed in a future release of `trialr`.

Furthermore, NBG propose a method for selecting dose that accounts for the probability of recommending an overdose.
That logic is currently not implemented in `escalation`, but will be in the near future.


## MCMC CRM
`escalation` already provides one-parameter CRM models via the `dfcrm` package [@dfcrm].
It now also provides MCMC implementations of those same one-parameter models:

```{r}
skeleton <- c(0.05, 0.12, 0.25, 0.40, 0.55)
target <- 0.25

model1 <- get_trialr_crm(skeleton = skeleton, target = target, 
                         model = 'empiric', beta_sd = 1)
model2 <- get_trialr_crm(skeleton = skeleton, target = target, 
                         model = 'logistic', a0 = 3, 
                         beta_mean = 0, beta_sd = 1)
```

and also adds a two-parameter logistic model:

```{r}
model3 <- get_trialr_crm(skeleton = skeleton, target = target, 
                         model = 'logistic2', 
                         alpha_mean = 0, alpha_sd = 2, 
                         beta_mean = 0, beta_sd = 1)
```

For more on the parameterisation of these methods, check out the CRM vignette in the [`escalation`](https://CRAN.R-project.org/package=escalation).

## Simulations and dose-paths
As with all methods in `escalation`, simulations are supported right out of the box.

For illustration, @Ji2007 present a simulation study to compare the performance of their TPI design to some alternatives.
They specify a design that uses a maximum sample size of thirty patients:

```{r}
model <- get_tpi(num_doses = 8, target = 0.25, k1 = 1, k2 = 1.5,
                 exclusion_certainty = 0.95) %>%
  stop_at_n(n = 30)
```

and their scenario 1 assumes true probability of toxicity:

```{r}
sc1 <- c(0.05, 0.25, 0.5, 0.6, 0.7, 0.8, 0.9, 0.95)
```

We can then reproduce their simulated characteristics by running:

```{r}
set.seed(123)
sims <- model %>%
  simulate_trials(num_sims = 50, true_prob_tox = sc1, next_dose = 1)

sims
```

For the sake of speed, we ran just fifty iterations but in a real situation, many thousands of iterations would be more appropriate.


## Conclusion
Amongst pure phase I designs, `escalation` now supports:

* CRM
* TPI
* mTPI
* BOIN
* NBG

It also supports some designs that select dose by co-primary efficacy and toxicity outcomes, the so-called seamless phase I/II designs.
Those will be the subect of the next post on this blog.
Til then, good day.



## References

